---
title: "ReadMe"
author: "Essi Shams"
date: "September 3, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This markdown file describes the steps taken in the run_analysis.R script to prepare and clean up the data. 
The first step taken is to load the dplyr library. This library is utilized heavily in this project. 

```{r}
library(dplyr)

```

The next step is setting the paths to the files that will need to be imported in order to load the data into the R environment. Each of the variables below hold the path to one of the text files in the current working directory.

```{r}
trainingFilePath = "UCI HAR Dataset/train/X_train.txt"
trainingLabelsFilePath = "UCI HAR Dataset/train/y_train.txt"
testFilePath ="UCI HAR Dataset/test/X_test.txt"
testLabelsFilePath = "UCI HAR Dataset/test/y_test.txt"
featureLabelsFilePath = "UCI HAR Dataset/features.txt"
trainSubjectsFilePath = "UCI HAR Dataset/train/subject_train.txt"
testSubjectsFilePath = "UCI HAR Dataset/test/subject_test.txt"
activitiesFilePath = "UCI HAR Dataset/activity_labels.txt"
```


the below code reads in the contents of the text files into the above data frames:

```{r}
trainingData <- read.table(trainingFilePath, sep="", header=FALSE, stringsAsFactors = FALSE)
trainingLabels <- read.table(trainingLabelsFilePath, sep="", header=FALSE, stringsAsFactors = FALSE)
testData <- read.table(testFilePath, sep="", header=FALSE, stringsAsFactors = FALSE)
testLabels <- read.table(testLabelsFilePath, sep="", header=FALSE, stringsAsFactors = FALSE)
featureLabels <- read.table(featureLabelsFilePath, sep="", header=FALSE, stringsAsFactors = FALSE)
testSubjects <- read.table(testSubjectsFilePath, sep="", header=FALSE, stringsAsFactors = FALSE)
trainingSubjects <- read.table(trainSubjectsFilePath, sep="", header=FALSE, stringsAsFactors = FALSE)
activities <- read.table(activitiesFilePath, sep="", header=FALSE, stringsAsFactors = FALSE)
```

For clarity and tidiness of data, column names are set for the dataframes:

```{r}
names(trainingSubjects) <- "subject"
names(testSubjects) <- "subject"
names(trainingLabels) <- "activityCode"
names(testLabels) <- "activityCode"
names(featureLabels) <- c("rowID","featureName")
featureLabels <- mutate(featureLabels,fullFeatureName=paste(rowID,featureName))
names(trainingData) <- featureLabels$fullFeatureName
names(testData) <- featureLabels$fullFeatureName
names(activities) <- c("activityCode", "activityDesc")
```

The trainingLabels and testLabels dataframes that hold the activity codes for each observation, are merged with the activties dataframe to obtain the activity description for each observation:

```{r}
trainingLabels <- merge(trainingLabels, activities, by.x="activityCode", by.y = "activityCode", sort = FALSE)
testLabels <- merge(testLabels, activities, by.x="activityCode", by.y = "activityCode", sort = FALSE)
```

Only the activityDesc column is retained from each of the above two dataframes, because we are only interested in having a description of the activity:

```{r}
trainingLabels <- select(trainingLabels,activityDesc)
testLabels <- select(testLabels, activityDesc)
```

because we are eventually going to combine our training and test data together, it is useful to have a column to identify whether each observation belonged to training or test data. 

```{r}
trainingData <- mutate(trainingData,obsType = "training")
testData <- mutate(testData, obsType = "test")
```

now we are ready to append the column containing the activity description, and the subject code to our training and test data respectively:

```{r}
trainingData <- cbind(trainingLabels, trainingData)
testData <- cbind(testLabels, testData)

trainingData <- cbind(trainingSubjects, trainingData)
testData <- cbind(testSubjects, testData)
```

and finally we combine both training and test datasets into one dataframe:

```{r}
allData <- rbind(trainingData, testData)
```

But we only want to retain the features that are either a mean or a std measurement.

```{r}
allData <- select(allData, subject, activityDesc, obsType, grep("mean|std",names(allData)))
```

We then group the data by subject and activity, and summarize each column to get its mean by the groups:

```{r}
groupedData <- group_by(allData, subject, obsType, activityDesc)
featureMeansBySubjectActivity <- summarize_each(groupedData,funs(mean))
```

